---
title: Tibbles and readr
---

A `tibble` is the modern `tidyverse` version of a `data.frame`. A `tibble` is a `data.frame`, and so can be used in the same way. But it comes with more powerful features and removes inconsistent and confusing behaviour.

In the same way, `readr` provides modern `tidyverse` replacements  for R's standard reading functions. `readr` provides `read_csv`, which is a better way of reading csv files than R's standard `read.csv`.

Let's now use the `tidyverse` to `read_csv` the dataset [Anatomical Data from Domestic Cats](https://vincentarelbundock.github.io/Rdatasets/doc/MASS/cats.html)  into a `tibble`. The first thing you will notice is that the `tidyverse` has printed out some useful information.

```{r}
#| echo: false
#| output: false
library(tidyverse)
```

```{r}
# Did you load tidyverse? library(tidyverse)

cats <- read_csv("https://raw.githubusercontent.com/Bristol-Training/intermediate-r/refs/heads/main/data/cats.csv")
```

This is telling you that `read_csv` found three columns; `Sex`, which is treated as a columns of strings (characters), and `BodyWeight` and `HeartWeight`, which are both treated as columns of floating point numbers (doubles).

Next, if you type `cats` and press return you will see

```{r}
cats
```


that the `tibble` summarises itself to the screen. This makes it much easier to quickly look at some data without it overflowing your console.

As a `tibble` is a `data.frame`, you can use the same methods of accessing data, e.g.

```{r}
cats$BodyWeight
```

```{r}
cats[1,]
```



::: {#exerciseN .callout-note icon=false title='Exercise'}

Load the cats data set into a `tibble` using `read_csv`. Use the `mean` and `max` functions to calculate the mean and max body weight and heart weight of the cats.

Note that according to [this dataset description](https://vincentarelbundock.github.io/Rdatasets/doc/MASS/cats.html) the body weight is in kilograms and the heart weight is in grams. As is normal for real-world data, there is little consistency in units.
:::

::: {#answerN .callout-caution icon=false title='Answer' collapse="true"}
```{R}
library(tidyverse)

cats <- read_csv("https://raw.githubusercontent.com/Bristol-Training/intermediate-r/refs/heads/main/data/cats.csv")

mean_body_weight <- mean(cats$BodyWeight)
max_body_weight <- max(cats$BodyWeight)
cat( sprintf("Body weight: mean = %.2f kg, max = %.2f kg\n",
             mean_body_weight, max_body_weight) )

cat( "Heart Weight: mean = ",
      mean(cats$HeartWeight) %>% round(digits=2),
      "g, max =",
      max(cats$HeartWeight) %>% round(digits=2),
      "g\n")

```
:::

